---
title: "iLOCO: Distribution-Free Inference for Feature Interactions"
author: "Samuel Switz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## iLOCO

The iLOCO metric measures pairwise feature interactions by comparing the effect of removing features individually vs jointly:

$$\text{iLOCO}_{j,k} = \Delta_j + \Delta_k - \Delta_{j,k}$$

where $\Delta_j$ is the change in prediction error when feature $j$ is removed.

### Data Setup (Scenario i from paper)
```{r}
set.seed(327)
M <- 10
N <- 500
B <- c(rnorm(5, 2, sqrt(0.5)), rep(0, 5))

X <- matrix(rnorm(M*N), nrow = N, ncol = M)

snr <- 2

# True interaction between features 1 and 2
f_X <- snr*(X[,1] * X[,2]) + X %*% B
e <- rnorm(N)
Y <- f_X + e
```

### iLOCO via Data Splitting (Algorithm 1)

```{r}
#' Compute iLOCO metric via data splitting
#'
#' @param X Feature matrix (N x M)
#' @param Y Response vector (N x 1)
#' @param j First feature index for interaction
#' @param k Second feature index for interaction
#' @param train_frac Fraction of data for training (default 0.5)
#' @param fit_model Function to fit model: function(X, Y) -> model
#' @param predict_model Function to predict: function(model, X) -> predictions
#' @param error_fn Error function: function(y_true, y_pred) -> error (default MSE)
#' @return List with iLOCO estimate, confidence interval, and component deltas
iLOCO_split <- function(X, Y, j, k, train_frac = 0.5,
                        fit_model = NULL, predict_model = NULL,
                        error_fn = function(y, yhat) (y - yhat)^2,
                        alpha = 0.1) {

  N <- nrow(X)
  M <- ncol(X)

  # Default to linear regression if no model specified
  if (is.null(fit_model)) {
    fit_model <- function(X, Y) lm(Y ~ ., data = data.frame(Y = Y, X))
  }
  if (is.null(predict_model)) {
    predict_model <- function(model, X) predict(model, newdata = data.frame(X))
  }

  # Split data
  train_idx <- sample(1:N, size = floor(train_frac * N))
  test_idx <- setdiff(1:N, train_idx)

  X_train <- X[train_idx, , drop = FALSE]
  Y_train <- Y[train_idx]
  X_test <- X[test_idx, , drop = FALSE]
  Y_test <- Y[test_idx]

  # Train full model
  f_full <- fit_model(X_train, Y_train)

  # Train model without feature j
  f_minus_j <- fit_model(X_train[, -j, drop = FALSE], Y_train)

  # Train model without feature k
  f_minus_k <- fit_model(X_train[, -k, drop = FALSE], Y_train)

  # Train model without features j and k
  f_minus_jk <- fit_model(X_train[, -c(j, k), drop = FALSE], Y_train)

  # Compute errors on test set
  err_full <- error_fn(Y_test, predict_model(f_full, X_test))
  err_minus_j <- error_fn(Y_test, predict_model(f_minus_j, X_test[, -j, drop = FALSE]))
  err_minus_k <- error_fn(Y_test, predict_model(f_minus_k, X_test[, -k, drop = FALSE]))
  err_minus_jk <- error_fn(Y_test, predict_model(f_minus_jk, X_test[, -c(j, k), drop = FALSE]))

  # Compute deltas for each test sample
  delta_j <- err_minus_j - err_full
  delta_k <- err_minus_k - err_full
  delta_jk <- err_minus_jk - err_full

  # iLOCO for each test sample
  iLOCO_samples <- delta_j + delta_k - delta_jk

  # Point estimate and confidence interval
  N_test <- length(test_idx)
  iLOCO_mean <- mean(iLOCO_samples)
  iLOCO_se <- sd(iLOCO_samples) / sqrt(N_test)
  z_alpha <- qnorm(1 - alpha/2)

  CI <- c(iLOCO_mean - z_alpha * iLOCO_se,
          iLOCO_mean + z_alpha * iLOCO_se)

  list(
    iLOCO = iLOCO_mean,
    CI = CI,
    se = iLOCO_se,
    delta_j = mean(delta_j),
    delta_k = mean(delta_k),
    delta_jk = mean(delta_jk),
    samples = iLOCO_samples
  )
}
```

### Compute iLOCO for all feature pairs

```{r}
#' Compute iLOCO for all unique feature pairs
#'
#' @param X Feature matrix
#' @param Y Response vector
#' @param ... Additional arguments passed to iLOCO_split
#' @return Data frame with iLOCO scores for all pairs
iLOCO_all_pairs <- function(X, Y, ...) {
  M <- ncol(X)
  pairs <- combn(1:M, 2)

  results <- data.frame(
    j = integer(),
    k = integer(),
    iLOCO = numeric(),
    CI_lower = numeric(),
    CI_upper = numeric(),
    delta_j = numeric(),
    delta_k = numeric(),
    delta_jk = numeric()
  )

  for (i in 1:ncol(pairs)) {
    j <- pairs[1, i]
    k <- pairs[2, i]

    res <- iLOCO_split(X, Y, j, k, ...)

    results <- rbind(results, data.frame(
      j = j,
      k = k,
      iLOCO = res$iLOCO,
      CI_lower = res$CI[1],
      CI_upper = res$CI[2],
      delta_j = res$delta_j,
      delta_k = res$delta_k,
      delta_jk = res$delta_jk
    ))
  }

  results[order(-results$iLOCO), ]
}
```

### Test on simulated data

The true interaction is between features 1 and 2 (snr * X1 * X2).

```{r}
# Compute iLOCO for the true interacting pair (1,2)
result_12 <- iLOCO_split(X, Y, j = 1, k = 2)
cat("iLOCO(1,2):", result_12$iLOCO, "\n")
cat("95% CI:", result_12$CI, "\n")
cat("Delta_1:", result_12$delta_j, "\n")
cat("Delta_2:", result_12$delta_k, "\n")
cat("Delta_12:", result_12$delta_jk, "\n")
```

```{r}
# Compute iLOCO for a null pair (no interaction)
result_34 <- iLOCO_split(X, Y, j = 3, k = 4)
cat("\niLOCO(3,4):", result_34$iLOCO, "\n")
cat("95% CI:", result_34$CI, "\n")
```

### Compute for all pairs and rank

```{r}
set.seed(327)  # For reproducibility of split
all_pairs <- iLOCO_all_pairs(X, Y)
head(all_pairs, 10)
```

### Visualization

```{r}
library(ggplot2)

# Add significance indicator
all_pairs$significant <- all_pairs$CI_lower > 0

# Create pair labels
all_pairs$pair <- paste0("(", all_pairs$j, ",", all_pairs$k, ")")

# Plot top 15 interactions
top_pairs <- head(all_pairs, 15)
top_pairs$pair <- factor(top_pairs$pair, levels = rev(top_pairs$pair))

ggplot(top_pairs, aes(x = iLOCO, y = pair)) +
  geom_point(aes(color = significant), size = 3) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper), height = 0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "blue") +
  scale_color_manual(values = c("FALSE" = "gray50", "TRUE" = "red")) +
  labs(x = "iLOCO Score", y = "Feature Pair",
       title = "Top 15 Feature Interactions by iLOCO",
       subtitle = "Red = CI excludes zero (significant)") +
  theme_minimal() +
  theme(legend.position = "none")
```

### Success probability simulation

Check if iLOCO correctly identifies (1,2) as the top interaction across multiple runs.

```{r}
simulate_success <- function(n_reps = 50, snr = 2) {
  successes <- 0

  for (rep in 1:n_reps) {
    # Generate new data
    X_sim <- matrix(rnorm(M*N), nrow = N, ncol = M)
    f_X_sim <- snr*(X_sim[,1] * X_sim[,2]) + X_sim %*% B
    Y_sim <- f_X_sim + rnorm(N)

    # Compute iLOCO for all pairs
    all_pairs_sim <- iLOCO_all_pairs(X_sim, Y_sim)

    # Check if (1,2) is ranked first
    top_pair <- all_pairs_sim[1, c("j", "k")]
    if ((top_pair$j == 1 && top_pair$k == 2) ||
        (top_pair$j == 2 && top_pair$k == 1)) {
      successes <- successes + 1
    }
  }

  successes / n_reps
}

# Run simulation (takes a few minutes)
success_prob <- simulate_success(n_reps = 50)
cat("Success probability:", success_prob, "\n")
```

## iLOCO via Minipatches (Algorithm 2)

The minipatch approach is more computationally efficient - after fitting the minipatch ensemble,
computing iLOCO for any pair is nearly free (no refitting needed).

```{r}
#' Fit minipatch ensemble
#'
#' @param X Feature matrix (N x M)
#' @param Y Response vector
#' @param B Number of minipatches
#' @param n_frac Fraction of observations per minipatch
#' @param m_frac Fraction of features per minipatch
#' @param fit_model Base learner fitting function
#' @return List with minipatch models and indices
fit_minipatch_ensemble <- function(X, Y, B = 1000, n_frac = 0.2, m_frac = 0.2,
                                    fit_model = NULL) {
  N <- nrow(X)
  M <- ncol(X)
  n <- max(2, floor(n_frac * N))  # observations per minipatch
  m <- max(2, floor(m_frac * M))  # features per minipatch

  if (is.null(fit_model)) {
    fit_model <- function(X, Y) lm(Y ~ ., data = data.frame(Y = Y, X))
  }

  models <- vector("list", B)
  obs_idx <- vector("list", B)   # I_b: observation indices
  feat_idx <- vector("list", B)  # F_b: feature indices

  for (b in 1:B) {
    # Sample observations and features
    I_b <- sample(1:N, n, replace = FALSE)
    F_b <- sample(1:M, m, replace = FALSE)

    # Fit model on minipatch
    X_b <- X[I_b, F_b, drop = FALSE]
    Y_b <- Y[I_b]

    models[[b]] <- tryCatch(
      fit_model(X_b, Y_b),
      error = function(e) NULL
    )
    obs_idx[[b]] <- I_b
    feat_idx[[b]] <- F_b
  }

  list(
    models = models,
    obs_idx = obs_idx,
    feat_idx = feat_idx,
    N = N,
    M = M,
    B = B
  )
}

#' Compute leave-one-out predictions from minipatch ensemble (vectorized)
#'
#' @param mp_ensemble Minipatch ensemble from fit_minipatch_ensemble
#' @param X Feature matrix
#' @param exclude_features Features to exclude (for LOCO prediction)
#' @param predict_model Prediction function
#' @return Matrix of LOO predictions (N x 1)
predict_minipatch_loo <- function(mp_ensemble, X, exclude_features = NULL,
                                   predict_model = NULL) {
  if (is.null(predict_model)) {
    predict_model <- function(model, X) predict(model, newdata = data.frame(X))
  }

  N <- mp_ensemble$N
  B <- mp_ensemble$B
  predictions <- rep(0, N)
  counts <- rep(0, N)

  for (b in 1:B) {
    model <- mp_ensemble$models[[b]]
    if (is.null(model)) next

    I_b <- mp_ensemble$obs_idx[[b]]
    F_b <- mp_ensemble$feat_idx[[b]]

    # Check if excluded features are NOT in this minipatch's feature set
    if (!is.null(exclude_features)) {
      if (any(exclude_features %in% F_b)) next
    }

    # For observations NOT in I_b (leave-one-out) - vectorized prediction
    left_out <- setdiff(1:N, I_b)

    # Predict all left-out observations at once
    preds <- tryCatch(
      predict_model(model, X[left_out, F_b, drop = FALSE]),
      error = function(e) rep(NA, length(left_out))
    )

    # Accumulate predictions
    valid <- !is.na(preds)
    predictions[left_out[valid]] <- predictions[left_out[valid]] + preds[valid]
    counts[left_out[valid]] <- counts[left_out[valid]] + 1
  }

  # Average predictions
  predictions / pmax(counts, 1)
}

#' Compute iLOCO using minipatch ensemble
#'
#' @param mp_ensemble Minipatch ensemble
#' @param X Feature matrix
#' @param Y Response vector
#' @param j First feature index
#' @param k Second feature index
#' @param error_fn Error function
#' @param alpha Significance level for CI
#' @return List with iLOCO estimate and CI
iLOCO_MP <- function(mp_ensemble, X, Y, j, k,
                     predict_model = NULL,
                     error_fn = function(y, yhat) (y - yhat)^2,
                     alpha = 0.1) {

  if (is.null(predict_model)) {
    predict_model <- function(model, X) predict(model, newdata = data.frame(X))
  }

  N <- nrow(X)

  # Get LOO predictions for each model variant
  pred_full <- predict_minipatch_loo(mp_ensemble, X, exclude_features = NULL, predict_model)
  pred_minus_j <- predict_minipatch_loo(mp_ensemble, X, exclude_features = j, predict_model)
  pred_minus_k <- predict_minipatch_loo(mp_ensemble, X, exclude_features = k, predict_model)
  pred_minus_jk <- predict_minipatch_loo(mp_ensemble, X, exclude_features = c(j, k), predict_model)

  # Compute errors
  err_full <- error_fn(Y, pred_full)
  err_minus_j <- error_fn(Y, pred_minus_j)
  err_minus_k <- error_fn(Y, pred_minus_k)
  err_minus_jk <- error_fn(Y, pred_minus_jk)

  # Compute deltas
  delta_j <- err_minus_j - err_full
  delta_k <- err_minus_k - err_full
  delta_jk <- err_minus_jk - err_full

  # iLOCO for each sample
  iLOCO_samples <- delta_j + delta_k - delta_jk

  # Point estimate and CI
  iLOCO_mean <- mean(iLOCO_samples, na.rm = TRUE)
  iLOCO_se <- sd(iLOCO_samples, na.rm = TRUE) / sqrt(sum(!is.na(iLOCO_samples)))
  z_alpha <- qnorm(1 - alpha/2)

  CI <- c(iLOCO_mean - z_alpha * iLOCO_se,
          iLOCO_mean + z_alpha * iLOCO_se)

  list(
    iLOCO = iLOCO_mean,
    CI = CI,
    se = iLOCO_se,
    delta_j = mean(delta_j, na.rm = TRUE),
    delta_k = mean(delta_k, na.rm = TRUE),
    delta_jk = mean(delta_jk, na.rm = TRUE),
    samples = iLOCO_samples
  )
}
```

### Test iLOCO-MP

```{r}
# Fit minipatch ensemble (this is done once)
# Note: B=500 is faster for demos; paper uses B=10,000+ for stable inference
set.seed(327)
cat("Fitting minipatch ensemble...\n")
mp_ensemble <- fit_minipatch_ensemble(X, Y, B = 500, n_frac = 0.2, m_frac = 0.5)
cat("Done! Fitted", mp_ensemble$B, "minipatches\n")

# Compute iLOCO-MP for the true interaction
cat("\nComputing iLOCO-MP predictions...\n")
result_mp_12 <- iLOCO_MP(mp_ensemble, X, Y, j = 1, k = 2)
cat("iLOCO-MP(1,2):", result_mp_12$iLOCO, "\n")
cat("90% CI:", result_mp_12$CI, "\n")

# Compute for a null pair
result_mp_34 <- iLOCO_MP(mp_ensemble, X, Y, j = 3, k = 4)
cat("\niLOCO-MP(3,4):", result_mp_34$iLOCO, "\n")
cat("90% CI:", result_mp_34$CI, "\n")
```

### Summary: iLOCO-Split vs iLOCO-MP

**iLOCO-Split** (recommended for most uses):
- Simple train/test split approach
- Trains 4 models per pair: full, -j, -k, -(j,k)
- Easy to understand and implement
- Works with any model class

**iLOCO-MP** (from paper):
- Fits minipatch ensemble once
- After fitting, iLOCO computation requires no additional model fitting
- Computationally efficient for large-scale analysis
- Better for production use with many pairs

For this demo, **iLOCO-Split is faster and simpler**. The minipatch method shines when:
1. You have many feature pairs to evaluate
2. You're using expensive base learners (neural nets, etc.)
3. You want to avoid data splitting

```{r}
# Compare the two methods on the true interaction pair
cat("Method comparison for pair (1,2):\n")
cat("iLOCO-Split:", result_12$iLOCO, "\n")
cat("iLOCO-MP:   ", result_mp_12$iLOCO, "\n")
```

### Interpretation of Results

A **positive iLOCO** value indicates the features j and k interact to improve predictions
beyond their individual effects.

A **negative iLOCO** can indicate correlated features that compensate for each other
(see Section 2.5 of the paper).

```{r}
# Show interpretation for our results
cat("Results interpretation:\n\n")
cat("Pair (1,2) - TRUE interaction:\n")
cat("  iLOCO =", round(result_12$iLOCO, 4), "\n")
cat("  CI excludes 0:", result_12$CI[1] > 0, "(significant)\n\n")

cat("Pair (3,4) - NULL (no interaction):\n")
cat("  iLOCO =", round(result_34$iLOCO, 4), "\n")
cat("  CI contains 0:", result_34$CI[1] <= 0 && result_34$CI[2] >= 0, "(not significant)\n")
```

## Higher-Order iLOCO (S-way interactions)

For S-way interactions, the iLOCO metric generalizes to:

$$\text{iLOCO}_S = \sum_{T \subseteq S, T \neq \emptyset} (-1)^{|T|+1} \Delta_T$$

```{r}
#' Compute higher-order iLOCO via data splitting
#'
#' @param X Feature matrix
#' @param Y Response vector
#' @param S Vector of feature indices for the interaction
#' @param train_frac Training fraction
#' @return List with iLOCO estimate
iLOCO_higher_order <- function(X, Y, S, train_frac = 0.5,
                                fit_model = NULL, predict_model = NULL,
                                error_fn = function(y, yhat) (y - yhat)^2) {

  if (is.null(fit_model)) {
    fit_model <- function(X, Y) lm(Y ~ ., data = data.frame(Y = Y, X))
  }
  if (is.null(predict_model)) {
    predict_model <- function(model, X) predict(model, newdata = data.frame(X))
  }

  N <- nrow(X)

  # Split data
  train_idx <- sample(1:N, size = floor(train_frac * N))
  test_idx <- setdiff(1:N, train_idx)

  X_train <- X[train_idx, , drop = FALSE]
  Y_train <- Y[train_idx]
  X_test <- X[test_idx, , drop = FALSE]
  Y_test <- Y[test_idx]

  # Train full model and compute baseline error
  f_full <- fit_model(X_train, Y_train)
  err_full <- error_fn(Y_test, predict_model(f_full, X_test))

  # Generate all non-empty subsets of S
  n_S <- length(S)
  subsets <- list()
  for (size in 1:n_S) {
    combos <- combn(S, size, simplify = FALSE)
    subsets <- c(subsets, combos)
  }

  # Compute Delta_T for each subset T
  iLOCO_samples <- rep(0, length(test_idx))

  for (T in subsets) {
    # Train model without features in T
    f_minus_T <- fit_model(X_train[, -T, drop = FALSE], Y_train)
    err_minus_T <- error_fn(Y_test, predict_model(f_minus_T, X_test[, -T, drop = FALSE]))

    delta_T <- err_minus_T - err_full

    # Add to iLOCO with alternating sign
    sign <- (-1)^(length(T) + 1)
    iLOCO_samples <- iLOCO_samples + sign * delta_T
  }

  list(
    iLOCO = mean(iLOCO_samples),
    se = sd(iLOCO_samples) / sqrt(length(test_idx)),
    samples = iLOCO_samples
  )
}
```

### Test third-order interaction

```{r}
# Generate data with a third-order interaction (Scenario iii)
set.seed(42)
X_3way <- matrix(rnorm(M*N), nrow = N, ncol = M)
f_X_3way <- snr * (X_3way[,1] * X_3way[,2] * X_3way[,3]) + X_3way %*% B
Y_3way <- f_X_3way + rnorm(N)

# Test 2nd order iLOCO on the 3-way interaction data
result_2nd <- iLOCO_split(X_3way, Y_3way, j = 1, k = 2)
cat("2nd-order iLOCO(1,2) on 3-way data:", result_2nd$iLOCO, "\n")

# Test 3rd order iLOCO
result_3rd <- iLOCO_higher_order(X_3way, Y_3way, S = c(1, 2, 3))
cat("3rd-order iLOCO(1,2,3):", result_3rd$iLOCO, "\n")

# Compare to a null triplet
result_3rd_null <- iLOCO_higher_order(X_3way, Y_3way, S = c(4, 5, 6))
cat("3rd-order iLOCO(4,5,6) [null]:", result_3rd_null$iLOCO, "\n")
```







